Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-02-10 _tex/paper2.tex
Built site for gh-pages

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics, which are unlikely to be equally
weighted or opposing, can exert an uneven influence on judges'
perceptions, creating biases in their judgments and, ultimately,
introducing dependencies between stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, thereby
favoring texts with fewer errors but weaker arguments. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

@@

In aggregate, the excessive reliance on Thurstone's Case V assumptions
in the statistical analysis of CJ data can compromise the reliability of
CJ's trait estimates. This overreliance, coupled with the apparent
disconnect between CJ's approach to trait measurement and hypothesis
testing, may also undermine the validity of these estimates
\citep[pp.~2]{Perron_et_al_2015}. Thus, addressing these issues through
a more systematic and integrated approach could provide significant
statistical and measurement benefits. For this reason, the study employs
the structural approach to causal inference.

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed extension. In this scenario, researchers aim to determine: ``To
what extent do different teaching methods influence students' ability to
produce high-quality written texts?'' To investigate this, they design a
CJ experiment by randomly assigning students (individuals) to two
groups, each receiving a different teaching method. Judges then compare
pairs of students' written texts (stimuli), producing a dichotomous
outcome that reflects the relative quality of each text (trait). Based
on this setup, researchers could reformulate the research question as
the estimand: ``\emph{On average}, is there a difference in the ability
to produce high-quality written texts between the two groups of
students?''. Following current CJ practices, researchers would then rely
on estimates from the BTL model, or its transformations, to approximate
this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify various
estimands relevant to CJ inquiries, including the one from the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. Without identification,
\emph{consistency} --the property of an estimator whose estimates
converge to the ``true'' value of the estimand as the data size
approaches infinity \citep{Everitt_et_al_2010}-- cannot be achieved,
even with ``infinite'' and error-free data. As a result, deriving
meaningful insights from finite data also becomes unattainable
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

Thus, this section addresses the identified limitations by extending
Thurstone's general form (see Table~\ref{tbl-thurstone_cases}) through a
systematic and integrated approach based on the structural approach to
causal inference. Specifically, it leverages the capabilities of this
approach to formalize the combination of the core theoretical principles
outlined in Section~\ref{sec-thurstone_theory} with key assessment
design features relevant to CJ experiments, such as the selection of
judges, stimuli, and comparisons.

In addition to overcoming the identified limitations, the extended model
offers two key advantages. First, it clarifies the interactions among
all actors and processes involved in CJ experiments. Second, it shifts
the current comparative data analysis paradigm from passively accepting
the model assumptions to actively testing if those assumptions fit the
specific data under analysis. For more details on these assumptions,
refer to Section~\ref{sec-theory-issues}.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what the study calls the
\emph{conceptual-population model}. This model assumes that researchers
have access to a conceptual population of comparison data, meaning data
representing all repeated comparisons made by every available judge for
each pair of stimuli produced by each pair of individuals in the
population, thus its name. Conversely,
Section~\ref{sec-theory-theoretical_SC} integrates the assessment design
features into what the study calls the \emph{sample-comparison model}.
This model assumes a more realistic scenario where researchers only have
access to a sample of judges, individuals, stimuli, and comparisons from
the conceptual population.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, the study defines SCMs. SCMs are
formal mathematical models characterized by a set of \emph{endogenous}
variables \(V\), a set of \emph{exogenous} variables \(E\), and a set of
functions \(F\) \citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous
variables are those whose causal mechanisms a researcher chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[pp.~27,68]{Pearl_2009}. Lastly, the functions, referred to as
\emph{structural equations}, express the endogenous variables as
non-parametric functions of other endogenous and exogenous variables.
These functions use the symbol `\(:=\)' to denote the asymmetrical
causal dependence between variables and the symbol `\(\:\bot\:\)' to
represent \emph{d-separation}, a concept akin to (conditional)
independence.

Figure~\ref{fig-cj04} illustrates the preliminary integration of
theoretical principles into the conceptual-population model.
Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
motivating example. This SCM maps the relationship between the outcome
of the CJ experiment \((O_{abhijk})\) to several other variables. In
this model, the indices \(\{a,b\}\) represent the compared texts
(stimuli), with \(a \neq b\). The indices \(\{h,i\}\) label the students
who wrote the texts (individuals), where \(h = i\) is possible. The
index \(j\) denotes the judge, and \(k\) denotes the comparison number,
which accounts for experimental cases where judges perform multiple
comparisons of the same texts written by the same students, i.e., a
\emph{repeated measures design} \citep[pp.~366-376]{Lawson_2015}.
Finally, the SCM presents the preliminary set of structural equations
\(F = \{f_{O},f_{D},f_{PT}\}\), which define the non-parametric
dependencies among these variables.

Note that the SCM~\ref{fig-cj04_scm} incorporates not only the variables
described by Thurstone's theory, such as the discriminal difference
\((D_{abhijk})\) and the texts' \emph{true} discriminal processes
\((T_{ahjk}, T_{bijk})\) (refer to Section~\ref{sec-thurstone_theory}),
but also innovates by incorporating additional variables central to CJ
experiments, such as the texts' \emph{perceived} discriminal processes
\((PT_{ahjk}, PT_{bijk})\) and the judges' perceptions \((B_{kj})\).
Together with the outcome, these variables constitute the preliminary
set of endogenous variables
\(V = \{O_{abhijk}, D_{abhijk}, PT_{ahjk}, PT_{bijk}, T_{ahjk}, T_{bijk}, B_{kj}\}\).

By incorporating the texts' perceived discriminal processes into the
conceptual-population model, this study considers a crucial aspect of CJ
experiments that previous literature has overlooked. It recognizes that
the texts' true discriminal processes become an observable outcome only
through the judges' perceptions. In other words, researchers observe the
CJ outcome only when both variables create a \emph{pure interaction
effect}, meaning that neither variable has a direct effect on the
outcome, but their interaction produces an effect
\citep{Attia_et_al_2022}.

Conversely, based on the evidence presented in
Section~\ref{sec-theory-issue1b}, the study assumes that other relevant
stimulus characteristics influence the judges' perceptions, leading to
biases that resist cancellation and ultimately introduce dependencies
between the stimuli. To address this, the study incorporates the judges'
perceptions into the conceptual-population model, following the
recommendations of \citet{Andrich_1978} and \citet{Wainer_et_al_1978}.

@@

Figure~\ref{fig-cj04_dag} displays the DAG associated with
SCM~\ref{fig-cj04_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ experiments. It also distinguishes between observable
endogenous variables, such as the outcome \(O_{abhijk}\), and variables
considered latent according to the theory and practice of CJ, such as
the discriminal difference, the perceived discriminal processes, the
true discriminal processes, and the judges' perceptions.

It is easy to intuit that describing the CJ system by assigning an index
and a variable to each discriminal process could quickly become
intractable and cumbersome due to the large size of the DAG and the
repetition of its components. Therefore, to maintain generality, the
study groups pairs of discriminal processes with different indices into
matrices that contain these pairs. First, the study defines the vectors
\(U_{A} \in \mathbb{N}^{A}\), \(U_{I} \in \mathbb{N}^{I}\) containing
the indices of the stimuli and individuals, where:

\[
U_{A} = [1, \dots, a, \dots, b, \dots, A] ;\quad U_{I} = [1, \dots, a, \dots, b, \dots, I]
\]

Then, the study defines the matrices of stimuli and individual
comparisons, \(P \in \mathbb{N}^{m \times 2}\) and
\(R \in \mathbb{N}^{n \times 2}\), respectively. Here, each row of \(P\)
represents a pairing \(\{a,b\}\) of stimuli indices and each row of
\(R\) represents a pairing \(\{h,i\}\) of individual indices, i.e.,

\[
P = \{(a, b) \mid a,b \in U_{A}, a \neq b\} ;\quad R = \{(h, i) \mid h,i \in U_{I}\}
\]

Here, \(m\) and \(n\) represents the number of unique pairwise
comparisons between stimuli, and unique pairwise comparison per
individuals, which corresponds to the number of ways to choose two
elements from \(U_{A}\) and \(U_{I}\), typically given by:

\[
m = {A \choose 2} = \frac{A!}{2!(A-2)!} ;\quad n = {I \choose 2} = \frac{I!}{2!(I-2)!}
\]

@@

\section{Extending the BTL model}\label{sec-statistical}

>>>>

In the first scenario, the intricate aspects of multidimensional,
complex traits may introduce dependencies between the stimuli due to
certain judges' biases that resist cancellation. Research on text
quality suggests that when judges evaluate these traits, they often rely
on various intricate characteristics of the stimuli to form their
judgments
\citep{vanDaal_et_al_2016, Lesterhuis_2018, Chambers_et_al_2022}. These
additional relevant characteristics are unlikely to be equally weighted
or opposing. As a result, they may exert an uneven influence on judges'
perceptions, creating biases in their judgments. Furthermore, since the
discriminal difference of the stimuli becomes an observable outcome only
through the judges' perceptions, these biases may ultimately introduce
dependencies between the stimuli
\citep[pp.~346]{vanderLinden_et_al_2017_II}. For example, this could
occur when a judge assessing the argumentative quality of a text places
more weight on its grammatical accuracy than other judges, thereby
favoring texts with fewer errors but weaker arguments. While direct
evidence for this particular scenario is lacking, studies such as
\citet{Pollitt_et_al_2003} demonstrate the presence of such biases,
supporting the notion that the factors influencing pairwise comparisons
may not always cancel out.

@@

In aggregate, the excessive reliance on Thurstone's Case V assumptions
in the statistical analysis of comparative data can compromise the
reliability of the trait estimates. This overreliance may also undermine
their validity \citep[pp.~2]{Perron_et_al_2015}, especially when
combined with the disconnect between CJ's approach to trait measurement
and hypothesis testing. However, the structural approach to causal
inference can address these issues by providing a systematic and
integrated framework that enhances statistical accuracy while
strengthening measurement reliability and validity.

@@

A motivating example helps to clarify these concepts. This example will
appear throughout the document to illustrate various aspects of the
proposed development. In this example, researchers aim to determine:
``To what extent do different teaching methods influence students'
ability to produce high-quality written texts?'' To investigate this,
researchers could design a CJ experiment by randomly assigning students
(individuals) to two groups, each receiving a different teaching method.
Judges would then compare pairs of students' written texts (stimuli), to
produce a dichotomous outcome that reflects the relative quality of each
text (trait). Based on this setup, researchers could reformulate the
research question as the estimand: ``\emph{On average}, is there a
difference in the ability to produce high-quality written texts between
the two groups of students?''. Finally, following current CJ practices,
researchers would then rely on estimates from the BTL model, or its
transformations, to approximate this estimand.

However, Section~\ref{sec-theory-issues} presents compelling evidence
that the BTL model has several statistical and measurement limitations.
These limitations hinder the model's ability to identify various
estimands relevant to CJ inquiries, including the one described in the
motivating example. Identification is crucial because it is a necessary
condition for ensuring consistent estimators. \emph{Consistency} refers
to the property of an estimator whose estimates converge to the ``true''
value of the estimand as the data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency cannot
be achieved, even with ``infinite'' and error-free data. Consequently,
deriving meaningful insights from finite data becomes impossible
\citep[pp.~5]{Schuessler_et_al_2023}.

@@

Thus, this section addresses the identified issues in
Section~\ref{sec-theory-issues} by extending Thurstone's general form
using the structural approach to causal inference. Specifically, it
leverages the capabilities of this approach to formalize the combination
of the core theoretical principles outlined in
Section~\ref{sec-thurstone_theory} with key assessment design features
relevant to CJ experiments, such as the selection of judges, stimuli,
and comparisons. In addition to enhancing statistical accuracy and
strengthening measurement reliability and validity, the approach offers
two key advantages. First, it clarifies the interactions among all
actors and processes involved in CJ experiments. Second, it shifts the
current comparative data analysis paradigm from passively accepting the
model assumptions to actively testing whether those assumptions fit the
data under analysis.

Accordingly, Section~\ref{sec-theory-theoretical_P} incorporates the
theoretical principles into what we call the \emph{conceptual-population
model}. This model assumes that researchers have access to a conceptual
population of comparison data, that is, data representing all repeated
comparisons made by every available judge for each pair of stimuli
produced by each pair of individuals in the population, hence its name.
Conversely, Section~\ref{sec-theory-theoretical_SC} integrates the
assessment design features into what we call the \emph{sample-comparison
model}. This model assumes a more realistic scenario where researchers
only have access to a sample of judges, individuals, stimuli, and
comparisons from the conceptual population.

@@

Before incorporating the theoretical principles into the
\emph{conceptual-population model}, it is essential to define SCMs. SCMs
are formal mathematical models characterized by a set of
\emph{endogenous} variables \(V\), a set of \emph{exogenous} variables
\(E\), and a set of functions \(F\)
\citep{Pearl_2009, Cinelli_et_al_2020}. Endogenous variables are those
whose causal mechanisms a researcher chooses to model \citep{Neal_2020}.
In contrast, exogenous variables represent \emph{errors} or
\emph{disturbances} arising from omitted factors that the investigator
chooses not to model explicitly \citep[pp.~27,68]{Pearl_2009}. Lastly,
the functions, referred to as \emph{structural equations}, express the
endogenous variables as non-parametric functions of other endogenous and
exogenous variables. These functions use the symbol `\(:=\)' to denote
the asymmetrical causal dependence between variables and the symbol
`\(\:\bot\:\)' to represent \emph{d-separation}, a concept akin to
(conditional) independence.

Figure~\ref{fig-cj04} illustrates the preliminary integration of
theoretical principles into the conceptual-population model.
Specifically, Figure~\ref{fig-cj04_scm} shows the SCM that underlies the
motivating example. This SCM maps the relationship between the
conceptual-population outcome \((O^{cp}_{abhijk})\) to several other
variables. In this SCM, the subscripts \(\{a,b\}\) represent the
compared texts (stimuli), with \(a \neq b\). The indices \(\{h,i\}\)
label the students who wrote the texts (individuals), where \(h = i\) is
possible. Moreover, the index \(j\) denotes the judge, and \(k\) denotes
the judgment number, which accounts for experimental cases where judges
perform multiple judgments of the same texts written by the same
students, i.e., a \emph{repeated measures design}
\citep[pp.~366-376]{Lawson_2015}.

The other variables in SCM~\ref{fig-cj04_scm} include the texts'
discriminal processes \((T_{ahjk}, T_{bijk})\) and their discriminal
difference \((D_{abhijk})\), in line with Thurstone's theory (see
Section~\ref{sec-thurstone_theory}). Additionally, based on the evidence
presented in Section~\ref{sec-theory-issue1b} and the recommendations of
\citet{Andrich_1978} and \citet{Wainer_et_al_1978}, we incorporate the
judges' biases \((B_{kj})\) into the model. Together with the outcome,
these variables form the preliminary set of endogenous variables,
\(V = \{ O_{abhijk}, D_{abhijk}, T_{ahjk}, T_{bijk}, B_{kj} \}\).
Finally, the SCM shows the preliminary set of structural equations,
\(F = \{ f_{O}, f_{D} \}\), which define the non-parametric dependencies
among these variables.

By incorporating judges' biases into the conceptual-population model, we
assume that additional relevant characteristics of the stimuli are
unlikely to be equally weighted or opposing. Thus, these characteristics
may exert an uneven influence on judges' perceptions, creating biases in
their judgments. Furthermore, since the discriminal difference of the
stimuli becomes an observable outcome only through the judges'
perceptions, these biases may ultimately introduce dependencies between
the stimuli \citep[pp.~346]{vanderLinden_et_al_2017_II}, hence the need
for their integration. The full statistical elaboration of this point is
presented in Section~\ref{sec-statistical}.

@@

Figure~\ref{fig-cj04_dag} displays the DAG associated with
SCM~\ref{fig-cj03_scm}. The graph shows the direction of causal
influences among variables, reflecting the assumptions of Thurstone's
theory and CJ experiments. It also distinguishes between observable
endogenous variables, such as the outcome, and variables considered
latent according to the theory and practice of CJ, such as the the
text's discriminal processes, their discriminal difference, and the
judges' biases.

It is easy to see that representing the CJ system using indices for a
single pair of individuals and stimuli, along with the indices for the
judgment number and the judge responsible for the comparison, can
quickly become intractable and cumbersome. Thus, to maintain generality
and simplify the notation, certain indices are grouped into vectors or
matrices. First, we define the matrices \(IA\) and \(JK\) as:

\[
IA = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
h & a \\
\vdots & \vdots \\
i & b \\
\vdots & \vdots \\
n_{I} & n_{A}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{I} \cdot n_{A}) \times 2} \; ; \quad
JK = \begin{bmatrix}
1 & 1 \\
1 & 2 \\
\vdots & \vdots \\
j & k \\
\vdots & \vdots \\
n_{J} & n_{K}-1 \\
n_{I} & n_{A} \\
\end{bmatrix} \in \mathbb{N}^{(n_{J} \cdot n_{K}) \times 2}
\]

Each row of \(IA\) represents a unique pairing of an individual
\(\{h,i\}\) with a stimulus \(\{a,b\}\). Here, \(n_{I}\) represents the
number of individuals and \(n_{A}\) the number of stimulus within
individuals in the population. Similarly, each row of \(JK\) represents
a unique pairing of a judge \(j\) with a judgment \(k\). Here, \(n_{J}\)
represents the number of judges and \(n_{K}\) the number of judgments
performed by each judge in the population.

@@

\section{Abandoning the BTL model}\label{sec-statistical}
